---
title: "Tutorial 11 - Kruskal-Wallis Test"
output: html_document
author: "Peter Mahoney, Andrej Panjkov"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Part 1 - Comparing averages

_
A psychologist wanted to test whether regular physical exercise had an effect on depression levels. Twenty four randomly selected people suffering from extreme depression were assigned to three treatment groups: subjects in Group A performed no physical exercise; subjects in Group B each exercised vigorously for 30 minutes per day; and subjects in Group C each exercised vigorously for 60 minutes per day. At the end of four weeks all subjects in the trial were asked to rate their level of depression on a scale from 1 (extremely depressed) to 100 (extremely happy). Their responses are shown below...
_

-----------------------------
 Group A   Group B   Group C
--------- --------- ---------
23            22	 59

26           27      66

51           39      38

49           29      49

58            46     39

37            48     59

29            59      46

44            55      63

--------- --------- ---------

```{r}
x <- t(matrix(c(23,22,59,
26,27,66,
51,39,38,
49,29,49,
58,46,39,
37,48,59,
29,59,46,
44,55,63), nrow=3,ncol=8))# R fills by column, so we transpose

depr.df<- data.frame(x)

colnames(depr.df)<-c("A","B","C")

depr.df
```

_
You must assess whether exercise has an effect on depression levels based on these data. Perform your calculations by hand or using Microsoft Excel.
_

_
a)	On what scale were the data measured?
_

Ordinal.

_
b)	Are you able to use a parametric test with this data?
_

No, the relevant parametric test that we have covered - ANOVA - requires normally distributed ratio data.


_
c)	What statistical test will you use to analyse this data?
_

The non-parametric Kruskal-Wallis comparison of medians.

_
d)	What will your null hypothesis (H0) be for this analysis?
_

H0:  The median level of depression is the same for each group.

_
e)	What will your alternate hypothesis (H1) be for this analysis?
_

H1:  The median level of depression is the different across the groups.

_
f)	What are the assumptions of your chosen test?
_

 * At least three groups factored by an independent variable.
 * Dependent variable must be numeric (not nominal).
 * At least 5 observations per group.
 * Each groups must have 'similar' distributions (possibly displaced).
 

_
g)	Do the data satisfy these assumptions? (Give your evidence here.)
_

We have 3 groups of ordinal data, with 8 observations each, so the first three conditions are met.

To assess the fourth, we begin by plotting the distributions.

```{r}
library(plotrix) #provides parallel histogram plotter
blue<-rgb(0,0.45,0.7,1/4)
orange<-rgb(0.9,0.6,0,1/4)
green<-rgb(0.0,0.9,0.4,1/4)
colours<-c(blue,orange,green)

multhist(depr.df, col=colours, freq=FALSE,
         main="Depression Level Histograms",
         xlab="Depression Level (0-100)",
         ylab="Relative Frequency")
legend("topright", colnames(depr.df), fill=colours)

```

There are too few data to conclude one way or another.

Let's look at some q-q plots, these also indicate similarity of distributions.

```{r}
ngroups = length(colnames(depr.df))



for (row in seq(1,ngroups-1)){
    for (col in seq(row+1,ngroups)) {
        qqplot(depr.df[[row]], depr.df[[col]],
            xlab=colnames(depr.df)[row],
            ylab=colnames(depr.df)[col],
            main="Q-Q plot")
        # qqPlot(lm(depr.df[[row]]~ depr.df[[col]]),
        #     xlab=colnames(depr.df)[row],
        #     ylab=colnames(depr.df)[col],
        #     main="Q-Q plot of studentized residuals")
    }
}

```

I think (need to check this) that if the distributions are similar, then their quantiles should track linearly, so a Q-Q plot should be near a straight line.  This appears to be the case.

Moving on...

_
h)	Calculate the test statistic for this analysis.
_

First we'll use R's builtin:

```{r}
result = kruskal.test(depr.df)
result
```

The high p-value means this is an unexceptional result, the medians are not different.

Now let's do the rank sum test explicitly.

```{r}
depr.stack.df <- stack(depr.df)
ranks <- rank(depr.stack.df$values)

# Attach ranks to pooled data.
depr.stack.df <- cbind( depr.stack.df, ranks)

#This code should be more functional.

# Sum the ranks for each group
sum.a <- sum(depr.stack.df$ranks[depr.stack.df$ind=="A"])
sum.b <- sum(depr.stack.df$ranks[depr.stack.df$ind=="B"])
sum.c <- sum(depr.stack.df$ranks[depr.stack.df$ind=="C"])

#sum.a ; sum.b ; sum.c
count.a = length(depr.df$A)
count.b = length(depr.df$B)
count.c = length(depr.df$C)
count.pool = count.a+count.b+count.c
H = 12/count.pool/(count.pool+1)*(
    sum.a^2/count.a +
    sum.b^2/count.b +
    sum.c^2/count.c) - 
    3*(count.pool+1)
H
```

This is very close to the result from `kruskal.test`, but with enough of a difference to warrant investigation. As we discovered in the tutorial, this algorithm does not include a correction for ties, of which there are 4 (in pairs).  SPSS does, and returns the same H statistic as `kruskal.test`.  The correction for ties is described in wikipedia, and when implemented in Excel, matches the corrected statistic offered by SPSS and R.

Here is some (unoptimised) code for applying the correction:

```{r}
#Detect ties, and count them. 
#(O(n^2) code, rewrite for large datasets)

# note use of unique function to prevent overcounting
correction <- 0.0

for (r in unique(depr.stack.df$ranks)){
    # extract all matching ranks
    match.r <- depr.stack.df$ranks[depr.stack.df$ranks==r]
    match.length <- length(match.r)
    if ( match.length > 1 ) {
        correction <- correction + 
            match.length^3 - match.length
    }
}
correction <- 1- correction/(count.pool^3-count.pool)


H/correction
```

...and we recover the corrected statistic.

_
i)	What is the critical value of the test statistic for this test (from the critical table)?
_

```{r}
qchisq(0.95,2)
```

_
j)	Do you accept or reject your null hypothesis?
_

We accepted the null based on the p-value above. 

Here, $H_{\rm calc} < H_{\rm crit}$, so this also has us accept the null.

_
k)	What does this mean for your alternate hypothesis?
_

I spit on the alternative hypothesis.

_
l)	Report your results in the correct format.
_

There was no significant difference between the median level of depression in subjects performing differing amounts of exercise (Kruskal-Wallis test; H(2) = 4.50; p > 0.05) 

(The remainder of the tutorial comprised non-computational questions.)

---
